On Optimality of Finite Element Integration

### 2015-10-18 TOMS_Paper/assembly.tex
Add section on transf space completeness

\begin{Def}[Reduction loop]
A loop $L_i$ is said to be a reduction loop if in its body
\begin{enumerate}
\item $i$ appears only as an array index, and
\item for each augmented assignment statement $S$ (e.g., increments), arrays indexed by $i$ appear only on the right hand side of $S$.
\end{enumerate}
\end{Def}

\begin{Def}[Free space loop]
A loop $L_i$ is said to be a free space loop if its iterations can be executed in any arbitrary order; that is, no dependencies exist across different iterations. 
\end{Def}

Altogether, these definitions lead us to formalize the class of loop nests for which we seek optimality.

\begin{Def}[Finite element integration loop nest]
A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free space loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
\end{Def}

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free space loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions). The expression \texttt{F} implements the operator.

@@

\subsection{Transformation Space}
To synthesize optimal implementations we need:
\begin{itemize}
\item a characterization of the space of legal transformations for the class of loop nests considered impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In this section, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.

We start with introducing the fundamental notion of sharing.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\end{figure}

@@

\begin{Algo}[Optimal code motion for outer product loop nests]
\label{algo:outer-product-optimal}
\normalfont
Consider an outer product loop nest $\Lambda = [L_i, L_j]$, with the body of $L_j$ being an independent basic block. The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_j$; this exposes the space of all possible factorizations. Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products. We traverse the expression tree and determine the sharing factor of the two linear loops. We take the symbols depending on the loop with smallest sharing factor, say $L_i$, and we incrementally factorize them. Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression. The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$. The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting (see~\citeN{Luporini} for more details about generalized code motion).
\end{Algo}

@@

We observe, however, that for finite element integration loop nests the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\end{figure}

@@

We expand on these points in Section~\ref{sec:optimal-synthesis}. In order to address point 2, we conclude this section by providing a more refined definition of optimality.

@@

@@

Note that due to the memory constraints imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space in which the optimal is searched. We consider them high level operators defined on top of expression-restructuring transformations such as code motion, re-association, and factorization. Other classes of loop or expression transformations, for instance aimed at improving data locality or exposing parallelism, can obviously be neglected since they do not impact the operation count.

Finally, with Proposition~\ref{prop:optimal-approach}, we elaborate on the optimality of this approach.

>>>>

\begin{Def}[Reduction loop]
A loop $L_i$ is said to be a reduction loop if in its body
\begin{enumerate}
\item $i$ appears only as an array index, and
\item for each augmented assignment statement $S$ (e.g., an increment), arrays indexed by $i$ appear only on the right hand side of $S$.
\end{enumerate}
\end{Def}

\begin{Def}[Free order loop]
A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, no loop-carried dependence exists across two different iterations. 
\end{Def}

Altogether, these definitions allow us to formalize the class of loop nests for which we seek optimality.

\begin{Def}[Finite element integration loop nest]
\label{def:fem-loopnest}
A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
\end{Def}

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over the elements in the mesh), a reduction loop $L_i$ (along which numerical integration is performed), and an outer product loop nest $[L_j, L_k]$ (over test and trial functions). The expression \texttt{F} implements the operator.

@@

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.

\subsection{Transformation Space - Sharing Elimination}
\label{sec:sharing-elimination}
We start with introducing the fundamental notion of sharing.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}

@@

\begin{Algo}[Optimal code motion for outer product loop nests]
\label{algo:outer-product-optimal}
\normalfont
Consider an outer product loop nest $\Lambda = [L_i, L_j]$, with the body of $L_j$ being an independent basic block. The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_j$; this exposes the space of all possible factorizations. Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products. We traverse the expression tree and determine the sharing factor of the two linear loops. We take the symbols depending on the loop with smallest sharing factor, say $L_i$, and we incrementally factorize them. Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression. The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$. The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
\end{Algo}

@@

\subsection{Transformation Space - Introducing Pre-evaluation}
\label{sec:pre-evaluation}
For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\label{code:loopnest_rednored}\end{figure}

@@

We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we provide a more refined definition of optimality.

@@

\subsection{Completeness of the Transformation Space}
It is helpful to consider sharing elimination and pre-evaluation as high level operators defined on top of basic transformations such as code motion, expansion, and factorization. These two operators allow us to analyze all possible ways in which an expression can be factorized; that is, by grouping:
\begin{itemize}
\item Test or trial functions (sharing elimination in the outer product loop nest)
\item Integration-dependent terms (sharing elimination in the reduction loop)
\item Together test functions, trial functions, and integration-dependent terms (pre-evaluation)
\end{itemize}
Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.

Sharing elimination and pre-evaluation also aim to remove \textit{temporal redundancy}, which arises whenever the same operation with the same operands is performed more than once. The problem of removing temporal redundancy reduces to minimize the execution of invariant sub-expressions. This requires two transformations:
\begin{itemize}
 \item Generalized code motion (occurs in both sharing elimination and pre-evaluation). Generalized, as explained in~\citeN{Luporini}, means that invariant sub-expressions are sought with respect to all loops composing a loop nest (and not just the innermost, as in traditional loop-invariant code motion). Examples were provided in Figures~\ref{code:multi_loopnest} and~\ref{code:loopnest_rednored}.
\item Evaluation of reduction loops (occurs in pre-evaluation). An example was provided in Figure~\ref{code:loopnest_rednored}.
 \end{itemize}

Sharing elimination and pre-evaluation (and, as we will clarify, their composition) allow us to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.

@@

Finally, with Proposition~\ref{prop:optimal-approach}, we show the optimality of this approach.
